\MakeUppercase{The Structure of $k$-Planar Graphs}

### 2019-07-09 k-planar.tex
Final preparations before submission

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}

\usepackage[longnamesfirst,numbers,sort&compress]{natbib}
\usepackage[noabbrev,capitalise]{cleveref}
\allowdisplaybreaks
%\sloppy
%%%%%%%%%%
\makeatletter
\def\NAT@spacechar{~}
\makeatother
%%%%%%%%%%
\crefname{lem}{Lemma}{Lemmas}
\crefname{thm}{Theorem}{Theorems}
\crefname{cor}{Corollary}{Corollaries}
\crefname{conj}{Conjecture}{Conjectures}
\crefname{obs}{Observation}{Observations}
\crefname{prop}{Proposition}{Propositions}
\crefname{clm}{Claim}{Claims}
\crefname{openproblem}{Open Problem}{Open Problems}
\crefformat{equation}{(#2#1#3)}
\Crefformat{equation}{Equation #2(#1)#3}

@@

\newcommand{\note}[2]{{\color{red}[#1:~#2]}}

@@

@@

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, and $p$-centred colouring. %, and graph encoding. 
We generalise this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
\end{titlepage}

@@

\tableofcontents
\newpage
\section{Introduction}
\label{Introduction}

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs and other more general classes of graphs. 
% Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.\note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }
The main results of this paper generalise the following recent theorem of \citet{dujmovic.joret.ea:planar} to this setting, and are the first results of this type for non-proper-minor-closed graph classes.

@@

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example.  
Loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
\begin{wrapfigure}{r}{0.5\textwidth} 
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{wrapfigure}
treewidth to be generalised for planar graphs (with different constants). \cref{PlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} proved that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992), and \citet{dujmovic.esperet.ea:planar}  proved that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002).

@@

We actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.} and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

@@

The proof of \cref{AddDummy} is trivial: Given a $k$-planar graph $G'$, create a planar graph $G$ by adding a dummy vertex at each crossing point. For each edge $vw\in E(G')\setminus E(G)$ there is a path $P$ in $G$ between $v$ and $w$ of length at most $k+1$. Let $\PP$ be the set of such paths $P$. For each vertex $v$ of $G$, at most two paths in $\PP$ use $v$ as an internal vertex (since no original vertex of $G'$ is an internal vertex of a path in $\PP$). Thus $\PP$ is a $(k+1,2)$-shortcut system for $G$, such that $G'=G^\PP$.

@@

\cref{ShortcutProduct} leads to several other results of interest. Here is one example. The \emph{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, we use $\dist_G(v,w)$ to denote the length of a shortest path with endpoints $v$ and $w$.  If $v$ and $w$ are in different components of $G$ then we use the convention that $\dist_G(v,w)=\infty$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\PP$ for some $(k,2k\Delta^{k})$-shortcut system $\PP$; see \cref{PowerShortcut}. \cref{kPlanarBasic,ShortcutProduct} then imply:

@@

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, and $p$-centred colouring  \citep{micek:personal}, 
%, and graph encoding \citep{BGP}. 
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

@@

\cref{Examples} presents several examples of graph classes that can be obtained from a shortcut system applied to a planar graph, including graph powers, map graphs, string graphs, and nearest neighbour graphs. All of these results also apply, where instead of planar graphs, we consider graphs of bounded Euler genus.   All of the applications in \cref{Applications}  are applicable to these examples.

@@

While strong products enable an easy statement of the theorems in \cref{Introduction}, to prove such results it is helpful to work with layerings and partitions, which we now introduce. A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in V_i$ and $w\in V_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.\footnote{An alternative definition of a layering of $G$ is an $H$-partition of $G$ where $H$ is a path.}  A layered $H$-partition $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{L},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

\note{DW}{I  prefer to not define ``layered $H$-partition''. Just define the ``layered width'' of an $H$-partition, which is what we did in \citep{dujmovic.joret.ea:planar}. There are many places where we write ``$H$-partition with layered width'' without writing ``layered $H$-partition''. I prefer it this way.}

\subsection{Previous Work on Partitions of Minor-Closed Classes}
\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has bounded treewidth. They defined a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition with layered width at most $\ell$ such that $H$ has treewidth at most $t$.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactenum}[(a)]
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.
\end{compactenum}
\end{thm}
Their proof is constructive and gives a simple quadratic-time algorithm for finding these partitions.

%Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define.

% \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs, graphs of bounded genus, and (more generally) apex-minor-free graphs.

% In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth:

\citet{dujmovic.joret.ea:planar} proved the following generalisation of \cref{PlanarBasic,PlanarPartition} for graphs embeddable on other surfaces.\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.}

@@

\begin{thm}\label{modern-major-general}
  Let $G$ be a graph having a layered $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\PP$ be a $(k,d)$-shortcut system for $G$.  Then $G^\PP$ has a layered $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ with treewidth at most $\binom{k+t}{t}-1$.
\end{thm}

@@

The relationship between $k$-planar graphs and $(k+1,2)$-shortcut systems of planar graphs shows that \cref{modern-major-general} has immediate applications to $k$-planar graphs.  The following result shows that $k$-planar graphs admit bounded layered partitions. With \cref{PartitionProduct,MakeWidth1}, this proves \cref{kPlanarBasic}. % with $c_k\leq(4+o(1))k^5$.

\begin{thm}
\label{k-planar}
Every $k$-planar graph has a layered $H$-partition of layered width at most $18k^2 + 30k$ in which $H$ has treewidth at most $\binom{k+4}{3}-1$. %$\treewidth$.
\end{thm}

\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with the weaker bound of $(2\times 3)((k+1)^3+3(k+1))=6k^3 + 18k^2 + 36k+24$ on the layered width. 
\note{DW}{I would just write, ``\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width.}
We prove the stronger bound in \cref{sec-k-planar}. 
%Also note that \cref{k-planar,MakeWidth1} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised. 
% PM: I removed the last line because it's already stated in the line preceding the theorem

@@

\begin{thm}
\label{1-planar}
Every 1-planar graph has a layered $H$-partition of layered width at most 30 where $H$ is planar and has treewidth at most 3.
\end{thm}

@@

The definition of $k$-planar graphs naturally generalises for other surfaces. A graph $G$ drawn on a surface $\Sigma$ is $(\Sigma,k)$-plane if every edge of $G$ is involved in at most $k$ crossings.  A graph $G$ is $(g,k)$-planar if it is isomorphic to some $(\Sigma,k)$-plane graph, for some surface $\Sigma$ with Euler genus at most $g$. 
 %\citet[Theorem~22]{dujmovic.joret.ea:planar} show that every genus-$g$ graph has a layered $H$-partition of layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most 4. 
\cref{AddDummy} immediately generalises as follows:

\begin{obs}
\label{gAddDummy}
Every $(g,k)$-planar graph is isomorphic to $G^\PP$ for some graph $G$ of Euler genus at most $g$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

\begin{thm}
\label{g-k-planar}
Every $(g,k)$-planar graph has a layered $H$-partition of layered width at most 
$\max\{2g,3\}\cdot(6k^2 + 10k)$ in which $H$ has treewidth at most $\binom{k+5}{4}-1$.
\end{thm}

Note that \cref{SurfacePartition}(b) and \cref{modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width. We prove the stronger bound in \cref{sec-k-planar}.

Finally, we state the following corollary of \cref{ApexMinorFree,modern-major-general,MakeWidth1,PartitionProduct}. This is the strongest result that follows from our work and that of \citet{dujmovic.joret.ea:planar}.

@@

The purpose of this section is to prove our main result, \cref{modern-major-general}. This theorem shows how, given a $(k,d)$-shortcut system $\mathcal{P}$ of a graph $G$, a layered $H$-partition of $G$ can be used to obtain a layered $J$-partition of $G^{\mathcal{P}}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.  Our main results for $k$-planar graphs (\cref{k-planar}) and $(g,k)$-planar graphs (\cref{g-k-planar}) follow; see \cref{sec-k-planar}.

@@

For the remainder of this section, $G$ is a graph and $(\mathcal{L}:=\langle V_1,\ldots,V_h\rangle,\mathcal{Y}:=(Y_x: x\in V(H)))$ is a layered partition of $G$ of layered width $\ell$ where the treewidth of the quotient graph $H=G/\mathcal{Y}$ is at most $t$.  In addition, $\mathcal{T}:=(B_x:x\in V(T))$ is a tree-decomposition of $H$ obtained by applying \lemref{nice-decomposition} to $H$.  The following lemma shows how to interpret the pair $(\mathcal{Y},\mathcal{T})$ as a hierarchical decomposition of $G$; refer to \cref{fig:generalized-tripod}.

@@

% Together, \clmref{general-width} and \clmref{general-bag-size} imply the following generalization of \thmref{k-planar}:
We are now ready to prove the main result of the paper.

>>>>

\documentclass{patmorin}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{amsmath,wrapfig}
\usepackage{amsfonts}
\usepackage{amsthm}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{amsfonts}
\usepackage{amsthm,mathtools}
\usepackage{pat}
\usepackage{paralist}
\usepackage{stmaryrd}
\usepackage{thm-restate}

\usepackage[longnamesfirst,numbers,sort&compress]{natbib}
\usepackage[noabbrev,capitalise]{cleveref}

\usepackage{doi} % To make plainnat handle doi's properly

\allowdisplaybreaks
%\sloppy
%%%%%%%%%%
\makeatletter
\def\NAT@spacechar{~}
\makeatother
%%%%%%%%%%
\crefname{lem}{Lemma}{Lemmas}
\crefname{thm}{Theorem}{Theorems}
\crefname{cor}{Corollary}{Corollaries}
\crefname{conj}{Conjecture}{Conjectures}
\crefname{obs}{Observation}{Observations}
\crefname{prop}{Proposition}{Propositions}
\crefname{clm}{Claim}{Claims}
\crefname{openproblem}{Open Problem}{Open Problems}
\crefformat{equation}{(#2#1#3)}
\Crefformat{equation}{Equation #2(#1)#3}

@@

% \newcommand{\note}[2]{{\color{red}[#1:~#2]}}
% \newcommand{\notex}[2]{{\color{blue}[#1:~#2]}}
\newcommand{\notex}[2]{}

@@

% \newcommand*{\doi}[1]{\href{https://dx.doi.org/#1}{\verb+doi: #1+}}

@@

\begin{abstract}
Dujmovi\'c~et~al.~(FOCS 2019) recently proved that every planar graph is a subgraph of the strong product of a graph of bounded treewidth and a path. This tool has been used to solve longstanding problems on queue layouts, non-repetitive colouring, $p$-centred colouring, and implicit graph encoding. We generalise this result for $k$-planar graphs, where a graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. In particular, we prove that every $k$-planar graph is a subgraph of the strong product of a graph of treewidth $O(k^5)$ and a path. This is the first result of this type for a non-minor-closed class of graphs. It implies, amongst other results, that $k$-planar graphs have non-repetitive chromatic number upper-bounded by a function of $k$. All these results generalise for drawings of graphs on arbitrary surfaces. In fact, we work in a much more general setting based on so-called shortcut systems that are of independent interest. 
\end{abstract}
\end{titlepage}

@@

% \tableofcontents
% \newpage

%\setcounter{\page}{1}
\section{Introduction}
\label{Introduction}

A graph is \emph{$k$-planar} if it has a drawing in the plane in which each edge is involved in at most $k$ crossings. Such graphs provide a natural generalisation of planar graphs, and are important in graph drawing research; see the recent bibliography on 1-planar graphs and the 140 references therein \citep{kobourov.liotta.ea:annotated}. The present paper studies the structure of $k$-planar graphs and other more general classes of graphs. 
\notex{PM}{I removed the following sentences because neither Vida nor could find anything about $k$-planar graphs in \cite{FP08} and, in any case, the proof is easy.  This could easily upset a referee.
% \note{PM}{I must be missing something.  The planarized graph $G'$ has $O(kn)$ vertices.  Let $S$ be the set of non-dummy vertices and take a separation $(A,B)$ of $G'$ with $|A\cap B|=O(\sqrt{kn})$, $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$.  Now take $X=A\cap B$ plus the endpoints of every edge of $G$ that contains a dummy vertex in $A\cap B$.  $|X|\le 5|A\cap B|=O(\sqrt{kn})$ and $X$ is a separator of $G$.} \note{DW}{I think you have just found a new proof of this fact. Note that you are actually using that $G'$ has treewidth $O(\sqrt{kn})$ and then applying the Robertson-Seymour separator lemma for graphs of given treewidth (so that $|S\cap (A\setminus B)|\le 2|S|/3$ and $|S\cap (B\setminus A)|\le 2|S|/3$). }
Various structural results about planar graphs generalise for $k$-planar graphs. For example, \citet{FP08} generalised the Lipton-Tarjan separator theorem to show that every $k$-planar graph with $n$ vertices has a balanced separator of order $O(\sqrt{kn})$.}
The main results of this paper generalise the following recent theorem of \citet{dujmovic.joret.ea:planar} to a setting that includes $k$-planar graphs.
 % as well as some other interesting non-minor-closed graph families.

@@

Here $\boxtimes$ is the strong product, and treewidth is an invariant that measures how `tree-like' a given graph is; see \cref{Tools} for formal definitions and see \cref{ProductExample} for an example. Loosely speaking, \cref{PlanarBasic} says that every planar graph is contained in the product of a tree-like graph and a path. This enables combinatorial results for graphs of bounded 
\begin{wrapfigure}{r}{0.5\textwidth} 
\centering
\includegraphics{ProductExample}
\caption{Example of a strong product.}
\label{ProductExample}
\end{wrapfigure}
treewidth to be generalised for planar graphs (with different constants). \cref{PlanarBasic} has been the key tool in solving several well-known open problems. In particular, \citet{dujmovic.joret.ea:planar} use it to prove that planar graphs have bounded queue-number (resolving a conjecture of \citet{HLR92} from 1992), \citet{dujmovic.esperet.ea:planar} use it to prove that planar graphs have bounded non-repetitive chromatic number (resolving a conjecture of \citet{AGHR-RSA02} from 2002), \citet{bonamy.gavoille.ea:shorter} use it to find shorter implicit representations of planar graphs.

@@

Although $k$-planar graphs are the most high-profile target for a generalization \cref{PlanarBasic}, we actually prove a substantially stronger result than \cref{kPlanarBasic} using the following definition. A collection $\mathcal{P}$ of paths in a graph $G$ is a \emph{$(k,d)$-shortcut system} (for $G$) if:
\begin{compactitem}
\item every path in $\mathcal{P}$ has length at most $k$,\footnote{A path of length $k$ consists of $k$ edges and $k+1$ vertices.} and
\item for every $v\in V(G)$, the number of paths in $\mathcal{P}$ that use $v$ as an internal vertex is at most $d$.
\end{compactitem} 
Each path $P\in\mathcal{P}$ is called a \emph{shortcut}; if $P$ has endpoints $v$ and $w$ then it is a \emph{$vw$-shortcut}. Given a graph $G$ and a $(k,d)$-shortcut system $\mathcal{P}$ for $G$, let $G^{\mathcal{P}}$ denote the supergraph of $G$ obtained by adding the edge $vw$ for each $vw$-shortcut in $\mathcal{P}$.

@@

The proof of \cref{AddDummy} is trivial: Given a $k$-plane embedding of a graph $G'$, create a planar graph $G$ by adding a dummy vertex at each crossing point. For each edge $vw\in E(G')$ there is a path $P$ in $G$ between $v$ and $w$ of length at most $k+1$. Let $\PP$ be the set of such paths $P$. For each vertex $v$ of $G$, at most two paths in $\PP$ use $v$ as an internal vertex (since no original vertex of $G'$ is an internal vertex of a path in $\PP$). Thus $\PP$ is a $(k+1,2)$-shortcut system for $G$, such that $G'\subseteq G^\PP$.

@@

\cref{ShortcutProduct} leads to several other results of interest. Here is one example. The \emph{$k$-th power} of a graph $G$ is the graph $G^k$ with vertex set $V(G^k):=V(G)$, where $vw\in E(G^k)$ if and only if $\dist_G(v,w)\leq k$.\footnote{For a graph $G$ and two vertices $v,w\in V(G)$, $\dist_G(v,w)$  denotes the length of a shortest path, in $G$, with endpoints $v$ and $w$.  We define $\dist_G(v,w):=\infty$ if $v$ and $w$ are in different connected components of $G$.} If $G$ has maximum degree $\Delta$, then $G^k = G^\PP$ for some $(k,2k\Delta^{k})$-shortcut system $\PP$; see \cref{PowerShortcut}. \cref{kPlanarBasic,ShortcutProduct} then imply:

@@

These theorems have applications in diverse areas, including queue layouts  \citep{dujmovic.joret.ea:planar}, non-repetitive colouring  \citep{dujmovic.esperet.ea:planar}, $p$-centred colouring  \citep{micek:personal}, and succinct graph representations \citep{bonamy.gavoille.ea:shorter}, some of which
%, and graph encoding \citep{BGP}. 
which we explore in \cref{Applications}. For example, we prove that $k$-planar graphs have bounded non-repetitive chromatic number (for fixed $k$). Prior to the recent work of \citet{dujmovic.esperet.ea:planar}, it was even open whether planar graphs have bounded non-repetitive chromatic number.

@@

\cref{Examples} presents several examples of graph classes that can be obtained from a shortcut system applied to a planar graph, including graph powers, map graphs, string graphs, and nearest neighbour graphs. All of these results also apply, where instead of planar graphs, we consider graphs of bounded Euler genus.   All of the applications discussed in \cref{Applications} work on these graph classes.

@@

While strong products enable concise statements of the theorems in \cref{Introduction}, to prove such results it is helpful to work with layerings and partitions, which we now introduce.

A \emph{layering} of a graph $G$ is a sequence $\mathcal{L}=\langle V_0,V_1,\ldots\rangle$ such that $\{V_0,V_1,\ldots\}$ is a partition of $V(G)$ and for every edge $vw\in E(G)$, if $v\in V_i$ and $w\in V_j$ then $|j-i|\leq 1$.  For any partition $\mathcal{P}=\{S_1,\ldots,S_p\}$ of $V(G)$, a \emph{quotient graph} $H=G/\mathcal{P}$ has a $p$-element vertex set $V(H)=\{x_1,\ldots,x_p\}$ and $x_ix_j\in E(H)$ if and only if there exists an edge $vw\in E(G)$ such that $v\in S_i$ and $w\in S_j$. To highlight the importance of the quotient graph $H$, we call $\mathcal{P}$ an \emph{$H$-partition} and write this concisely as $\mathcal{P}=\{S_x : x\in V(H)\}$ so that each element of $\mathcal{P}$ is indexed by the vertex it creates in $H$.

For any partition $\mathcal{P}$ of $V(G)$ and any layering $\mathcal{L}$ of $G$ we define the \emph{layered width} of $\mathcal{P}$ with respect to $\mathcal{L}$ as $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.  For any partition $\mathcal{P}$ of $V(G)$, we define the \emph{layered width} of $\mathcal{P}$ as the minimum, over all layerings $\mathcal{L}$ of $G$, of the layered width of $\mathcal{P}$ with respect to $\mathcal{L}$.

% \footnote{An alternative definition of a layering of $G$ is an $H$-partition of $G$ where $H$ is a path.}

% A \emph{layered $H$-partition} $(\mathcal{L},\mathcal{P})$ of a graph $G$ consists of a layering $\mathcal{L}$ of $G$ and an $H$-partition $\mathcal{P}$ of $G$. The \emph{layered width} of $(\mathcal{L},\mathcal{P})$ is $\max\{|L\cap P|: L\in\mathcal{L},\, P\in\mathcal{P}\}$.

% \note{DW}{I  prefer to not define ``layered $H$-partition''. Just define the ``layered width'' of an $H$-partition, which is what we did in \citep{dujmovic.joret.ea:planar}. There are many places where we write ``$H$-partition with layered width'' without writing ``layered $H$-partition''. I prefer it this way.}

\subsection{Previous Work on Partitions of Minor-Closed Classes}

\citet{dujmovic.joret.ea:planar} introduced the study of partitions with bounded layered width such that the quotient has some additional desirable property, like small treewidth. 
% Say that a graph $G$ admits a $(t,\ell)$-partition if $V(G)$ has a partition $\mathcal{P}$ of layered width at most $\ell$ and the treewidth of $H:=G/\mathcal{P}$ is at most $t$. 
They defined a class $\mathcal{G}$ of graphs to \emph{admit bounded layered partitions} if there exist $t,\ell\in\mathbb{N}$ such that every graph $G\in \mathcal{G}$ has an $H$-partition of layered width at most $\ell$ for some graph $H=H(G)$ of treewidth at most $t$.

@@

\begin{thm}[\citep{dujmovic.joret.ea:planar}]
\label{PlanarPartition}
Every planar graph has:
\begin{compactenum}[(a)]
\item an $H$-partition of layered width $1$ for some planar graph $H$ of treewidth at most $8$, and
\item an $H$-partition of layered width $3$ for some planar graph $H$ of treewidth at most $3$.
\end{compactenum}
\end{thm}
Their proof is constructive and gives a simple quadratic-time algorithm for finding these partitions and corresponding layerings.
%Layered $H$-partitions of small layered width in which $H$ has some additional property are a useful tool for a number of graph problems.  One such useful property is small \emph{treewidth}, which we now define. 
%
% \citet{dujmovic.joret.ea:planar} show that, if each graph $G$ in some family $\mathcal{G}$ of graphs has a layered $H$-partition for which the layered width of the partition and the treewidth of $H$ are each upper-bounded by some constant $c_\mathcal{G}$, then each of the following quantities are upper bounded by a constant, for every $G\in\mathcal{G}$: queue number, track number, and layered treewidth. \citet{dujmovic.esperet.ea:planar} add non-repetitive chromatic number to this list.  \citet{dujmovic.joret.ea:planar} show that two common families of graphs have such $H$-partitions: planar graphs, graphs of bounded genus, and (more generally) apex-minor-free graphs.
%
% In this paper we show that another family of graphs admits layered $H$-partitions of bounded layered width in which $H$ has bounded treewidth:
The same authors proved the following generalisation of \cref{PlanarBasic,PlanarPartition} for graphs embeddable on other surfaces.\footnote{The \textit{Euler genus} of the orientable surface with $h$ handles is $2h$. The \textit{Euler genus} of the non-orientable surface with $c$ cross-caps is $c$. The \textit{Euler genus} of a graph $G$ is the minimum integer $g$ such that $G$ embeds in a surface of Euler genus $g$. Of course, a graph is planar if and only if it has Euler genus 0; see \citep{mohar.thomassen:graphs} for more about graph embeddings in surfaces.}

@@

\begin{restatable}{thm}{mmg}
  %\begin{thm}
  \label{modern-major-general}
    Let $G$ be a graph having an $H$-partition of layered width $\ell$ in which $H$ has treewidth at most $t$ and let $\PP$ be a $(k,d)$-shortcut system for $G$.  Then $G^\PP$ has a $J$-partition of layered width at most $d\ell(k^3+3k)$ for some graph $J$ with treewidth at most $\binom{k+t}{t}-1$.
  %\end{thm}
\end{restatable}

@@

Using relationship between $k$-planar graphs and $(k+1,2)$-shortcut systems along with a direct application of \cref{modern-major-general} we can obtain a slightly weaker version of \cref{k-planar}, below. The minor modifications needed to obtain the stronger bound are described in \cref{sec-k-planar}.

 % of planar graphs shows that \cref{modern-major-general} has immediate applications to $k$-planar graphs.  The following result shows that $k$-planar graphs admit bounded layered partitions. With \cref{PartitionProduct,MakeWidth1}, this proves \cref{kPlanarBasic}. % with $c_k\leq(4+o(1))k^5$.

\begin{thm}
\label{k-planar}
Every $k$-planar graph has an $H$-partition of layered width at most $18k^2 + 30k$ in which $H$ has treewidth at most $\binom{k+4}{3}-1$. %$\treewidth$.
\end{thm}

% \cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with the weaker bound of $O(k^3)$
%  % $(2\times 3)((k+1)^3+3(k+1))=6k^3 + 18k^2 + 36k+24$
% on the layered width. 
% % \note{DW}{I would just write, ``\cref{PlanarPartition,modern-major-general} imply \cref{k-planar} with a weaker bound on the layered width.}
% We prove the stronger bound in \cref{sec-k-planar}. 
% %Also note that \cref{k-planar,MakeWidth1} imply \cref{kPlanarBasic} with the $O(k^5)$ bound on the treewidth of $H$, as promised. 
% % PM: I removed the last line because it's already stated in the line preceding the theorem

@@

\begin{thm}
\label{1-planar}
Every 1-planar graph has an $H$-partition of layered width at most 30 where $H$ is planar and has treewidth at most 3.
\end{thm}

@@

The definition of $k$-planar graphs naturally generalises for other surfaces. A graph $G$ drawn on a surface $\Sigma$ is $(\Sigma,k)$-plane if every edge of $G$ is involved in at most $k$ crossings.  A graph $G$ is $(g,k)$-planar if it is isomorphic to some $(\Sigma,k)$-plane graph, for some surface $\Sigma$ with Euler genus at most $g$. \citet{grigoriev.bodlaender:algorithms} study approximation algorithms for $(g,k)$-plane graphs.
 %\citet[Theorem~22]{dujmovic.joret.ea:planar} show that every genus-$g$ graph has a layered $H$-partition of layered width at most $\max\{2g,3\}$ such that $H$ has treewidth at most 4. 
\cref{AddDummy} immediately generalises as follows:

\begin{obs}
\label{gAddDummy}
Every $(g,k)$-planar graph is isomorphic to a subgraph of $G^\PP$ for some graph $G$ of Euler genus at most $g$ and some $(k+1,2)$-shortcut system $\PP$ for $G$. 
\end{obs}

@@

\begin{thm}
\label{g-k-planar}
Every $(g,k)$-planar graph has an $H$-partition of layered width at most 
$\max\{2g,3\}\cdot(6k^2 + 10k)$ in which $H$ has treewidth at most $\binom{k+5}{4}-1$.
\end{thm}

Again, a direct application of \cref{modern-major-general} using \cref{SurfacePartition}(b) implies \cref{k-planar} with a weaker bound on the layered width. We prove the stronger bound in \cref{sec-k-planar}.

Finally, we state the following corollary of \cref{ApexMinorFree,modern-major-general,MakeWidth1}. This is the most general result that follows from \cref{modern-major-general} and that of \citet{dujmovic.joret.ea:planar}.

@@

The purpose of this section is to prove our main result, \cref{modern-major-general}. This theorem shows how, given a $(k,d)$-shortcut system $\mathcal{P}$ of a graph $G$, a $H$-partition of $G$ can be used to obtain a $J$-partition of $G^{\mathcal{P}}$ where the layered width  does not increase dramatically and the treewidth of $J$ is not much more than the treewidth of $H$.  Our main results for $k$-planar graphs (\cref{k-planar}) and $(g,k)$-planar graphs (\cref{g-k-planar}) follow; see \cref{sec-k-planar}.

@@

@@

% Together, \clmref{general-width} and \clmref{general-bag-size} imply the following generalization of \thmref{k-planar}:
